% ============================================================================
% Botho Executive Summary - 2-Page Overview
% ============================================================================

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}

% Compact spacing
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0pt}

\title{\vspace{-2em}\textbf{Botho: Privacy-Preserving Cryptocurrency}\\[0.3em]
\large Executive Summary}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-3em}

\section*{Overview}

\textbf{Botho} is a privacy-preserving cryptocurrency that combines post-quantum
cryptography with efficient transaction privacy, achieving strong security
guarantees with practical performance.

\textbf{Key innovation}: Hybrid post-quantum architecture applies quantum-resistant
cryptography selectively---protecting permanent data (recipient identities) while
using efficient classical cryptography for ephemeral privacy (sender anonymity).

\section*{Core Features}

\begin{tabular}{@{}p{3.5cm}p{10cm}@{}}
\toprule
\textbf{Feature} & \textbf{Description} \\
\midrule
Privacy & All transactions private by default. Ring signatures (CLSAG) hide
sender among 20 decoys. Stealth addresses hide recipients. Confidential
transactions hide amounts. \\[0.5em]
Post-Quantum & ML-KEM-768 protects recipient identity against future quantum
attacks. On-chain data remains secure even if quantum computers break
classical cryptography. \\[0.5em]
Fast Finality & Hybrid PoW+SCP consensus achieves deterministic finality in
$\sim$5 seconds. Unlike Bitcoin, finalized blocks cannot be reverted. \\[0.5em]
Fair Economics & Progressive fees based on coin ancestry (not identity) create
Sybil-resistant pressure against wealth concentration. Fees are 80\%
redistributed via lottery, 20\% burned. \\
\bottomrule
\end{tabular}

\section*{Technical Parameters}

\begin{tabular}{@{}lcl@{}}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Notes} \\
\midrule
Ring size & 20 & Sender hidden among 20 possibilities \\
Transaction size & $\sim$4 KB & Practical for mobile wallets \\
Block time & 5--40s & Dynamic based on network load \\
Finality & $\sim$5s & Deterministic (not probabilistic) \\
Tail emission & 0.3 BTH/block & Perpetual $\sim$2\% inflation \\
\bottomrule
\end{tabular}

\section*{Cryptographic Primitives}

\begin{itemize}
\item \textbf{Recipient privacy}: ML-KEM-768 (post-quantum KEM, NIST FIPS 203)
\item \textbf{Sender privacy}: CLSAG ring signatures (classical, efficient)
\item \textbf{Amount privacy}: Pedersen commitments + Bulletproofs range proofs
\item \textbf{Minting signatures}: ML-DSA-65 (post-quantum signatures, NIST FIPS 204)
\item \textbf{Key derivation}: BIP39 mnemonic $\rightarrow$ SLIP-10 $\rightarrow$ subaddresses
\end{itemize}

\section*{Consensus Mechanism}

\textbf{Hybrid PoW + SCP}: Proof-of-work provides permissionless block proposal
(anyone can mine). Stellar Consensus Protocol provides fast Byzantine-fault-tolerant
finalization.

\textbf{Why hybrid?}
\begin{itemize}
\item PoW alone: Slow probabilistic finality, reorg vulnerability
\item BFT alone: Requires known validator set, not permissionless
\item PoW+SCP: Permissionless \emph{and} fast deterministic finality
\end{itemize}

\section*{Economic Design}

\textbf{Progressive fees}: Transaction fees scale with sender's cluster wealth
percentile (1$\times$ for bottom 50\%, up to 6$\times$ for top 1\%). Based on
coin \emph{ancestry}, not current ownership---preserves privacy.

\textbf{Fee distribution}:
\begin{itemize}
\item 80\% redistributed via lottery to random UTXO (favors small holders statistically)
\item 20\% burned (creates deflationary pressure proportional to usage)
\end{itemize}

\textbf{Tail emission}: Perpetual 0.3 BTH per block ensures long-term security
funding. Asymptotic inflation: $\sim$2\%.

\section*{Security Model}

\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Threat} & \textbf{Protection} & \textbf{Security Level} \\
\midrule
Sender identification & CLSAG ring signatures & 128-bit classical \\
Recipient identification & ML-KEM-768 stealth addresses & 192-bit post-quantum \\
Amount discovery & Pedersen + Bulletproofs & 128-bit classical \\
Double-spending & Key images + SCP finality & Information-theoretic \\
51\% attack & SCP quorum intersection & Byzantine fault tolerant \\
\bottomrule
\end{tabular}

\section*{Comparison}

\begin{tabular}{@{}lccccc@{}}
\toprule
& \textbf{Botho} & \textbf{Monero} & \textbf{Zcash} & \textbf{Bitcoin} \\
\midrule
Privacy & Full & Full & Optional & Pseudonymous \\
Post-quantum & Partial & No & No & No \\
Finality & 5s deterministic & 10+ min prob. & 10+ min prob. & 60+ min prob. \\
Tx size & 4 KB & 2 KB & 2 KB (shielded) & 0.3 KB \\
Fair distribution & Yes & No & No & No \\
\bottomrule
\end{tabular}

\section*{Resources}

\begin{itemize}
\item \textbf{Full whitepaper}: \texttt{botho-whitepaper.pdf} (127 pages)
\item \textbf{Website}: \url{https://botho.org}
\item \textbf{Source code}: \url{https://github.com/botho/botho}
\end{itemize}

\vfill
\begin{center}
\small\textit{``Motho ke motho ka batho'' --- A person is a person through other people.}
\end{center}

\end{document}
