% ============================================================================
% Section 4: Cryptographic Protocol
% ============================================================================

\section{Cryptographic Protocol}
\label{sec:cryptography}

\subsection{Key Hierarchy}

\Botho derives all cryptographic keys from a single BIP39~\cite{bip39}
mnemonic using SLIP-10~\cite{slip10} hierarchical derivation.

\subsubsection{Master Seed Generation}

A 24-word mnemonic encodes 256 bits of entropy. The master seed is derived
via:
\begin{equation}
\text{seed} = \text{PBKDF2-SHA512}(\text{mnemonic}, \text{``mnemonic''} \| \text{passphrase}, 2048)
\end{equation}

This produces a 512-bit seed from which all keys are derived.

\subsubsection{Account Key Derivation}

Using SLIP-10 hardened derivation with path $m/44'/866'/\text{account}'$:

\begin{align}
\text{view\_key\_material} &= \text{SLIP10}(\text{seed}, m/44'/866'/0'/0') \\
\text{spend\_key\_material} &= \text{SLIP10}(\text{seed}, m/44'/866'/0'/1')
\end{align}

These materials are then processed through HKDF-SHA512~\cite{rfc5869} to produce the
final Ristretto255 scalars:

\begin{align}
a &= \text{HKDF}(\text{view\_key\_material}, \text{``botho-ristretto255-view''}) \mod q \\
b &= \text{HKDF}(\text{spend\_key\_material}, \text{``botho-ristretto255-spend''}) \mod q
\end{align}

The corresponding public keys are:
\begin{equation}
A = aG, \quad B = bG
\end{equation}

The public address is the tuple $(A, B)$.

\input{figures/key-hierarchy}

\subsubsection{Subaddress Derivation}

Subaddresses allow generation of unlimited unlinkable addresses from a
single master key pair. \Botho uses the \emph{multiplicative} subaddress
construction (as in MobileCoin), which differs from Monero's additive
scheme: the view subaddress key is the product $a \cdot D_i$ rather than
a shifted point $A + \delta_i G$. For subaddress index $i$, a
domain-separated scalar is derived with Blake2b-512:

\begin{equation}
\delta_i = \Hs(\text{``bth\_subaddress''} \| a \| i),
\quad \Hs(\cdot) = \text{Blake2b-512}(\cdot) \bmod q
\end{equation}

The subaddress private key pair $(c_i, d_i)$ --- view and spend
respectively --- is then:

\begin{align}
d_i &= \delta_i + b \mod q \\
c_i &= a \cdot d_i = a\,(\delta_i + b) \mod q
\end{align}

with corresponding public keys:

\begin{equation}
D_i = d_i G = \delta_i G + B, \qquad C_i = c_i G = a \cdot D_i
\end{equation}

Here $D_i$ is the subaddress spend public key and $C_i$ is the
subaddress view public key; the published subaddress is the pair
$(C_i, D_i)$. Note that the view key is obtained multiplicatively as
$a \cdot D_i$, not by the additive offset $A + \delta_i G$: this is the
defining difference from the Monero construction and is what the
reference implementation computes.

\begin{theorem}[Subaddress Unlinkability]
Without knowledge of the view key $a$, subaddresses $(C_i, D_i)$ and
$(C_j, D_j)$ for $i \neq j$ are computationally indistinguishable from
independent random points.
\end{theorem}

\subsection{Post-Quantum Stealth Addresses}
\label{sec:pq-stealth}

Traditional CryptoNote stealth addresses use ECDH for key
exchange~\cite{cryptonote,courtois-stealth}, building on the dual-key
stealth-address construction (DKSAP). \Botho keeps that classical ECDH step
and \emph{additionally} binds an ML-KEM-768 encapsulation into the same
one-time key---a \emph{hybrid} construction. Each output's one-time key is
derived from BOTH the classical Diffie--Hellman shared secret and the ML-KEM
shared secret, so an attacker must break \emph{both} elliptic-curve ECDH
\emph{and} lattice ML-KEM to recover or link an output. This achieves
post-quantum recipient privacy without discarding the classical security the
network already relies on.

\paragraph{Intuition.}
Recipient privacy in a classical CryptoNote coin rests entirely on the
hardness of the ECDH problem: the link between an output and its owner is an
elliptic-curve Diffie--Hellman shared secret. That secret is not disclosed on
chain, but everything needed to \emph{recompute} it under a quantum attacker
is---the sender's ephemeral public key sits in the transaction permanently.
An adversary who records the chain today and later runs Shor's algorithm
recovers the shared secret and deanonymizes every historical recipient. This
is the \emph{harvest-now, decrypt-later} exposure, and because recipient
identity is permanent on-chain data (Table~\ref{tab:hybrid}) it cannot be
mitigated by waiting or rotating keys after the fact.

Folding an ML-KEM-768 encapsulation into the one-time key---alongside, not in
place of, the classical ECDH secret---closes this exposure at its source. The
sender encapsulates a fresh shared secret against the recipient's KEM public
key and mixes it with the classical Diffie--Hellman secret when deriving the
output's one-time key; recovering that key requires breaking a lattice problem
believed hard even for quantum adversaries, so the recipient-unlinkability
reduction of the Security Analysis below survives quantumly. Because the
derivation is hybrid, the output also remains at least as private as a classical
CryptoNote coin against any adversary who has \emph{not} broken ECDH: a flaw in
the lattice assumption alone cannot deanonymize the recipient. Crucially,
the KEM hardens only the \emph{key-exchange} step---the part whose secrecy
must last forever---while sender anonymity continues to rely on the compact
classical ring signature of Section~\ref{sec:clsag}. Making the entire spend
authorization post-quantum would instead demand lattice ring signatures, whose
$\sim$50$\times$ size overhead we quantify in
Section~\ref{sec:hybrid-rationale} under ``Why not full post-quantum?''; the
hybrid split buys quantum-durable recipient privacy for the cost of a single
1{,}088-byte ciphertext per output rather than that blowup.

\subsubsection{Protocol Description}

The recipient's ML-KEM keypair is \emph{not} derived from the Ristretto
address point. Instead it is generated independently from the same BIP39
seed as the rest of the account-key hierarchy: the 512-bit seed is fed
through HKDF-SHA256 with domain separator \texttt{"botho-pq-v1"} and info
label \texttt{"kem-seed"} to produce a 32-byte ML-KEM seed, from which the
ML-KEM-768 keypair $(\pk_{\text{kem}}, \sk_{\text{kem}})$ is deterministically
generated (a sibling \texttt{"sig-seed"} label likewise seeds the ML-DSA-65
signing keypair). Consequently the recipient's address publishes
$\pk_{\text{kem}}$ directly alongside the Ristretto spend key $B$; there is
no hash-of-point step, which would in any case not yield a valid ML-KEM key.

The one-time key binds \emph{two} shared secrets: the classical ECDH secret
$\dh = rC = aR$ (as in classical CryptoNote, where $r$ is the sender's
ephemeral scalar, $R = rG$ the published transaction public key, $C$ the
recipient's view key, and $a$ its secret) \emph{and} the ML-KEM shared secret
$K$. The stealth scalar is the domain-separated hash of both:
\[
  s = \Hs\bigl(\dh \,\|\, K \,\|\, \text{output\_index}\bigr).
\]

\textbf{Sender} (creating output for recipient):
\begin{enumerate}
  \item Obtain the recipient's view key $C$ and ML-KEM public key
    $\pk_{\text{kem}}$ from their address
  \item Sample ephemeral scalar $r$; publish transaction public key $R = rG$
  \item Compute classical DH secret: $\dh = rC$
  \item Encapsulate: $(c, K) \leftarrow \MLKEM.\Encap(\pk_{\text{kem}})$
  \item Compute hybrid scalar: $s = \Hs(\dh \| K \| \text{output\_index})$
  \item Compute one-time public key: $P = sG + B$
  \item Include ciphertext $c$ (1,088 bytes) in transaction output
\end{enumerate}

\textbf{Recipient} (scanning for received outputs):
\begin{enumerate}
  \item Regenerate the ML-KEM secret key $\sk_{\text{kem}}$ from the BIP39
    seed (HKDF label \texttt{"kem-seed"}, as above)
  \item For each output with transaction public key $R$ and ciphertext $c$:
    \begin{enumerate}
      \item Recompute classical DH secret: $\dh = aR$ (\,$= rC$\,)
      \item Decapsulate: $K \leftarrow \MLKEM.\Decap(\sk_{\text{kem}}, c)$
      \item Compute hybrid scalar: $s' = \Hs(\dh \| K \| \text{output\_index})$
      \item Compute expected key: $P' = s'G + B$
      \item If $P' = P$ (output's public key), this output belongs to us
      \item Compute private key: $x = s' + b \mod q$
    \end{enumerate}
\end{enumerate}

\textbf{Verification}: The one-time private key $x$ satisfies
$xG = (s + b)G = sG + B = P$, and because $aR = a(rG) = r(aG) = rC$ the sender
and recipient derive the identical hybrid scalar $s = s'$.

\subsubsection{Security Analysis}

\begin{theorem}[Recipient Unlinkability]
If \emph{either} the ML-KEM-768 encapsulation is IND-CCA2 secure \emph{or} the
elliptic-curve Diffie--Hellman problem is hard, an adversary cannot link
outputs to recipients with probability better than negligible, given only the
blockchain. In particular, a quantum adversary---who is assumed to break
ECDH---is still defeated by the ML-KEM branch, while an adversary who breaks
only the lattice assumption is still defeated by the classical branch.
\end{theorem}

\begin{proof}
Because the stealth scalar $s = \Hs(\dh \| K \| \text{output\_index})$ is a
domain-separated hash of \emph{both} shared secrets, modelling $\Hs$ as a
random oracle makes $s$ (hence the one-time key $P$) pseudorandom whenever
\emph{either} $\dh$ or $K$ is unknown to the adversary. It therefore suffices
to exhibit a reduction for each branch; we give the post-quantum (ML-KEM)
branch, as the classical ECDH branch is the standard CryptoNote
unlinkability argument~\cite{cryptonote} applied to the $\dh$ input.

We prove the ML-KEM branch by reduction to ML-KEM IND-CCA2 security. Suppose
adversary $\Adv$ can distinguish outputs belonging to recipient $R$ with
non-negligible advantage $\epsilon$, given $\dh$ (i.e.\ even after ECDH is
broken). We construct adversary $\Adv'$ that breaks ML-KEM IND-CCA2 with
advantage $\epsilon/2$.

\textbf{Setup}: $\Adv'$ receives ML-KEM public key $\pk^*$ from the
IND-CCA2 challenger. $\Adv'$ sets this as the target recipient's
KEM public key.

\textbf{Query phase}: $\Adv$ may request outputs for various recipients.
For non-target recipients, $\Adv'$ generates honestly. For the target
recipient, $\Adv'$ uses the decapsulation oracle.

\textbf{Challenge}: $\Adv$ submits two recipients $R_0, R_1$ (one being
the target). $\Adv'$ queries the IND-CCA2 challenger with
$(\pk^*, \pk_{\text{other}})$ and receives ciphertext $c^*$ encapsulating
either $K_0$ or $K_1$ (random bit $b$).

$\Adv'$ computes $s^* = \Hs(\dh \| K_b \| \text{index})$ and
$P^* = s^*G + B_b$, returning output $(c^*, P^*)$ to $\Adv$.

\textbf{Analysis}: If $\Adv$ correctly identifies the recipient with
probability $1/2 + \epsilon$, then $\Adv'$ correctly guesses $b$ with
the same probability, contradicting IND-CCA2 security.

The reduction is tight: $\text{Adv}^{\text{IND-CCA2}}_{\Adv'} \geq \epsilon/2$.
\end{proof}

\begin{corollary}[Quantum Security]
The above reduction holds against quantum adversaries since ML-KEM-768
provides IND-CCA2 security in the quantum random oracle model (QROM).
\end{corollary}

\input{figures/stealth-address}

\subsection{Ring Signatures (CLSAG)}
\label{sec:clsag}

\CLSAG (Concise Linkable Spontaneous Anonymous Group)~\cite{clsag}
provides efficient linkable ring signatures~\cite{fujisaki2007} for sender privacy.

\subsubsection{Ring Construction}

For each input being spent, the sender:
\begin{enumerate}
  \item Selects $n-1$ decoy outputs from the blockchain (we use $n = 20$)
  \item Forms a ring $\mathcal{R} = \{(P_0, C_0), \ldots, (P_{n-1}, C_{n-1})\}$
    where each $(P_i, C_i)$ is a one-time public key and commitment
  \item The real input is at secret index $\pi$
\end{enumerate}

\subsubsection{Signature Generation}

Given:
\begin{itemize}
  \item Ring $\mathcal{R}$ with real index $\pi$
  \item Private key $x_\pi$ and commitment blinding factor $z_\pi$
  \item Message $m$ (transaction hash)
\end{itemize}

\textbf{Key Image}: $I = x_\pi \cdot \Hp(P_\pi)$

The key image is deterministic given the private key and serves as a
unique tag preventing double-spending.

\textbf{Aggregation Coefficients}:
\begin{align}
\mu_P &= \Hash(\text{``CLSAG\_agg\_0''} \| \mathcal{R} \| I \| D) \\
\mu_C &= \Hash(\text{``CLSAG\_agg\_1''} \| \mathcal{R} \| I \| D)
\end{align}

where $D$ is an auxiliary point for commitment verification.

\textbf{Signature}: The signature $\sigma = (c_0, s_0, \ldots, s_{n-1}, I, D)$
is computed via a Fiat-Shamir transform~\cite{fiatshamir} of an interactive protocol,
forming a ``ring'' of challenges that closes only if the prover knows
one of the private keys.

\subsubsection{Verification}

Given signature $\sigma$ and ring $\mathcal{R}$:
\begin{enumerate}
  \item Recompute aggregation coefficients $\mu_P$, $\mu_C$
  \item For $i = 0, \ldots, n-1$:
    \begin{align}
      W_i &= \mu_P P_i + \mu_C (C_i - C_{\text{out}}) \\
      L_i &= s_i G + c_i W_i \\
      R_i &= s_i \Hp(P_i) + c_i (\mu_P I + \mu_C D) \\
      c_{i+1} &= \Hash(\text{``CLSAG\_round''} \| \mathcal{R} \| L_i \| R_i \| m)
    \end{align}
  \item Accept if $c_n = c_0$ (ring closure)
\end{enumerate}

\subsubsection{Security Properties}

\begin{definition}[Unforgeability Game]
The unforgeability experiment $\text{Exp}^{\text{UNF}}_{\Adv}$ proceeds:
\begin{enumerate}
  \item Challenger generates key pairs $(x_i, P_i)$ for $i \in [n]$
  \item $\Adv$ receives $\{P_i\}$ and access to signing oracle
  \item $\Adv$ outputs $(m^*, \sigma^*, \mathcal{R}^*)$
  \item $\Adv$ wins if $\Verify(\sigma^*, m^*, \mathcal{R}^*) = 1$ and
    $m^*$ was never queried to signing oracle with ring $\mathcal{R}^*$
\end{enumerate}
\end{definition}

\begin{definition}[Anonymity Game]
The anonymity experiment $\text{Exp}^{\text{ANON}}_{\Adv}$ proceeds:
\begin{enumerate}
  \item Challenger generates key pairs, gives public keys to $\Adv$
  \item $\Adv$ selects ring $\mathcal{R}$, message $m$, two indices $i_0, i_1$
  \item Challenger flips bit $b$, signs with key $x_{i_b}$
  \item $\Adv$ outputs guess $b'$
  \item $\Adv$'s advantage: $|\Pr[b' = b] - 1/2|$
\end{enumerate}
\end{definition}

\begin{theorem}[CLSAG Security]
Under the DLP assumption in the random oracle model, CLSAG satisfies:
\begin{enumerate}
  \item \textbf{Unforgeability}: For all PPT $\Adv$,
    $\Pr[\text{Exp}^{\text{UNF}}_{\Adv} = 1] \leq \negl(\lambda)$.
  \item \textbf{Anonymity}: For all PPT $\Adv$,
    $\text{Adv}^{\text{ANON}}_{\Adv} \leq \negl(\lambda)$.
  \item \textbf{Linkability}: For all PPT $\Adv$,
    $\Pr[\text{same key, different images}] \leq \negl(\lambda)$ and
    $\Pr[\text{different keys, same image}] \leq \negl(\lambda)$.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof (Unforgeability)]
By reduction to DLP. Given DLP instance $(G, Y = xG)$, embed $Y$ as one
ring member's public key. A forking lemma argument shows that a successful
forger can be rewound to extract the discrete log, contradicting DLP
hardness. Full proof in~\cite{clsag}.
\end{proof}

\begin{proof}[Proof (Anonymity)]
The signature is a Fiat-Shamir transform of a $\Sigma$-protocol. In the
random oracle model, the simulated transcript is indistinguishable from
real, regardless of which key was used. The ring structure ensures all
positions are computationally indistinguishable.
\end{proof}

\begin{proof}[Proof (Linkability)]
Key image $I = x \cdot \Hp(P)$ is deterministic given $(x, P)$. For the
same key, the same image is always produced. For different keys $x \neq x'$,
collision requires $x \cdot \Hp(P) = x' \cdot \Hp(P')$, which occurs with
probability $1/q$ (negligible) when $\Hp$ is modeled as a random oracle.
\end{proof}

\input{figures/ring-signature}

\subsection{Confidential Transactions}

\input{figures/confidential-tx}

\subsubsection{Amount Commitment}

Each output commits to its amount $v$ with blinding factor $r$:
\begin{equation}
C = vH + rG
\end{equation}

The blinding factor $r$ is derived deterministically from the shared
secret to enable recipient recovery.

\subsubsection{Value Conservation}

For a transaction with inputs $\{C_{\text{in}}^{(i)}\}$ and outputs
$\{C_{\text{out}}^{(j)}\}$, value conservation requires:
\begin{equation}
\sum_i C_{\text{in}}^{(i)} = \sum_j C_{\text{out}}^{(j)} + fH
\end{equation}

where $f$ is the transaction fee (public). This holds if and only if:
\begin{equation}
\sum_i v_{\text{in}}^{(i)} = \sum_j v_{\text{out}}^{(j)} + f
\end{equation}

Validators check commitment arithmetic without learning individual values.

\subsubsection{Range Proofs}

Each output includes a Bulletproof demonstrating:
\begin{equation}
v_{\text{out}}^{(j)} \in [0, 2^{64})
\end{equation}

This prevents:
\begin{itemize}
  \item Negative amounts (which would create coins from nothing)
  \item Overflow attacks (amounts wrapping around)
\end{itemize}

\subsection{Minting Attribution}
\label{sec:minting-attribution}

Block rewards (minting transactions) carry no signature. The minter's
identity is public, and attribution is bound by the proof of work
itself: the RandomX preimage commits to the minter's public address
keys,

\begin{equation}
h_{\text{PoW}} = \text{RandomX}\bigl(\text{nonce} \| \text{prev\_hash}
\| \pk_{\text{view}} \| \pk_{\text{spend}}\bigr)
\end{equation}

and the minting transaction is valid only if $h_{\text{PoW}}$ meets the
difficulty target. Reattributing a block's reward would change the
minting transaction's hash --- and with it the cluster origin identifier
and the stealth reward output derived from it
(Section~\ref{sec:cluster-tags}) --- and would require redoing the proof
of work against a chain whose blocks are already externalized by
consensus (Section~\ref{sec:consensus}). Because the binding is
hash-based, it is subject only to Grover-type quantum speedups and
remains sound against a cryptographically relevant quantum computer:
minting attribution needs no post-quantum signature, and adding one
(ML-DSA-65's 3{,}309-byte signature plus 1{,}952-byte public key on
every block) would cost roughly 33~GB of chain growth per year at the
reference block time for no additional security in this model.

\subsection{Hybrid Architecture Rationale}
\label{sec:hybrid-rationale}

\begin{table}[h]
\centering
\caption{Cryptographic choices by data lifetime}
\label{tab:hybrid}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Data} & \textbf{Lifetime} & \textbf{Algorithm} & \textbf{Rationale} \\
\midrule
Recipient identity & Permanent & ML-KEM-768 & On-chain forever; must be PQ \\
Sender identity & Ephemeral & CLSAG & Value degrades; efficiency wins \\
Amounts & Permanent & Pedersen & Information-theoretic hiding \\
Minting attribution & Permanent & PoW hash binding & Hash-based; quantum-resistant \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Why not full post-quantum?}

Post-quantum ring signatures (e.g., lattice-based constructions) impose
approximately 50$\times$ size overhead. A CLSAG signature is $\sim$700
bytes per input; a comparable lattice ring signature would be
$\sim$35 KB. With multiple inputs, transactions would exceed 100 KB,
making desktop nodes impractical.

\textbf{Why is ephemeral sender privacy acceptable?}

The value of sender deanonymization degrades over time. Learning who
sent a transaction in 2025 from a 2045 perspective has minimal economic
relevance---the goods have been delivered, contracts fulfilled, and
economic context forgotten. In contrast, recipient identity remains
valuable (``who owns this address?'') indefinitely.

This asymmetry justifies asymmetric protection: permanent data gets
permanent (post-quantum) protection; ephemeral data gets efficient
(classical) protection.
