% ============================================================================
% Section 7: Monetary Policy
% ============================================================================

\section{Monetary Policy}
\label{sec:monetary}

\Botho's monetary policy balances long-term security sustainability with
controlled inflation, using dynamic mechanisms that respond to network
conditions.

\subsection{Design Goals}

\begin{enumerate}
  \item \textbf{Sustainable security}: Perpetual miner incentive without
    relying solely on transaction fees.
  \item \textbf{Fair distribution}: Rewards distributed over time, not
    front-loaded to early adopters.
  \item \textbf{Anti-hoarding}: Economic pressure toward circulation rather
    than accumulation.
  \item \textbf{Predictability}: Clear, auditable emission schedule.
\end{enumerate}

\subsection{Emission Schedule}

\subsubsection{Initial Emission}

Block rewards follow a halving schedule:

\begin{equation}
R(h) = R_0 \cdot \left(\frac{1}{2}\right)^{\lfloor h/H \rfloor}
\end{equation}

where:
\begin{itemize}
  \item $R(h)$ is the block reward at height $h$
  \item $R_0 = 50$ BTH is the initial block reward
  \item $H = 6{,}307{,}200$ blocks ($\approx$ 1 year at 5s blocks)
    is the halving period
\end{itemize}

After $K = 5$ halvings ($\approx$ 5 years), the per-block reward has fallen
from 50 BTH to 3.125 BTH and the chain transitions to perpetual tail
emission. The geometric sum of the five halving epochs fixes the
\emph{Phase-1 supply} at
\begin{equation}
S_1 = R_0 H \sum_{i=0}^{K-1} 2^{-i}
    = 50 \times 6{,}307{,}200 \times \tfrac{31}{16}
    = 611{,}010{,}000 \text{ BTH} \;(\approx 611\text{M}).
\end{equation}

\input{figures/emission-curve}

\subsubsection{Tail Emission}

After the five halvings complete, perpetual tail emission begins. Rather than
a fixed per-block constant, the tail targets a constant \emph{net annual
inflation rate} on the circulating supply, so emission scales with supply
rather than being pinned to an absolute number:

\begin{equation}
R_{\text{tail}} = \frac{S \cdot (\beta_{\text{net}} + \beta_{\text{burn}})}
                       {N_{\text{year}}}
\end{equation}

where:
\begin{itemize}
  \item $S$ is the circulating supply at the start of the tail phase
  \item $\beta_{\text{net}} = 0.02$ (200 bps) is the target net annual
    inflation
  \item $\beta_{\text{burn}} = 0.005$ (50 bps) is the expected annual
    fee-burn rate (the 20\% burn, see Section~\ref{sec:economic-modeling}),
    added so that \emph{gross} emission nets to the 2\% target after burns
  \item $N_{\text{year}} = 6{,}307{,}200$ is the number of blocks per year at
    5s blocks
\end{itemize}

Anchoring this to the Phase-1 supply $S_1 = 611{,}010{,}000$ BTH gives, at
tail onset:
\begin{align}
\text{Gross annual emission} &= 611{,}010{,}000 \times 0.025 = 15{,}275{,}250 \text{ BTH} \\
\text{Net annual emission}   &= 611{,}010{,}000 \times 0.020 = 12{,}220{,}200 \text{ BTH} \\
R_{\text{tail}} &= \frac{15{,}275{,}250}{6{,}307{,}200} \approx 2.42 \text{ BTH/block (gross)}
\end{align}
or $\approx 1.94$ BTH/block net of the expected fee burn. Because $R_{\text{tail}}$
is proportional to supply, the per-block reward \emph{grows} as the supply
grows, holding net inflation fixed at 2\%.

\subsubsection{Supply Projection}

\begin{table}[h]
\centering
\caption{Supply projection}
\label{tab:supply}
\begin{tabular}{@{}rrrr@{}}
\toprule
\textbf{Year} & \textbf{Circulating} & \textbf{New Emission} & \textbf{Gross Inflation} \\
\midrule
1 & 315.4M BTH & 315.4M & --- \\
2 & 473.0M BTH & 157.7M & 50\% \\
3 & 551.9M BTH & 78.8M & 17\% \\
5 & 611.0M BTH & 19.7M & 3.3\% \\
6 & 626.3M BTH & 15.3M & 2.5\% \\
10 & 691.3M BTH & 16.9M & 2.5\% \\
20 & 884.9M BTH & 21.6M & 2.5\% \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Note}: Years 1--5 are the halving phase (rewards 50, 25, 12.5, 6.25,
3.125 BTH/block), which fix the Phase-1 supply at $\approx$611M BTH. From year 6
onward the tail holds \emph{gross} emission at 2.5\% of supply (2\% net of the
expected 0.5\% fee burn). Because the tail tracks supply, nominal emission
grows over time while the net inflation rate stays pinned at 2\%.

\input{figures/supply-projection}

\subsection{Dynamic Block Timing}
\label{sec:dynamic-timing}

\Botho introduces adaptive block intervals that respond to network
utilization, creating natural inflation dampening.

\subsubsection{Mechanism}

The target block time is a deterministic function of recent transaction
load, computed over a smoothing window of the last 10 blocks. Rather than a
continuous curve, the protocol selects from five discrete levels keyed on the
measured transaction rate (transactions per second):

\begin{equation}
T_{\text{target}}(\rho) =
\begin{cases}
3\text{s}  & \rho \geq 20 \text{ tx/s} \\
5\text{s}  & \rho \geq 5  \text{ tx/s} \\
10\text{s} & \rho \geq 1  \text{ tx/s} \\
20\text{s} & \rho \geq 0.2 \text{ tx/s} \\
40\text{s} & \text{otherwise (idle)}
\end{cases}
\end{equation}

where $\rho$ is the transaction rate measured over the smoothing window.
The absolute consensus floor is $T_{\min} = 3$ seconds (SCP needs time to
externalize) and the idle ceiling is $T_{\max} = 40$ seconds. Because the
level is computed from committed chain state, every validator derives the
same target for a given tip.

\begin{table}[h]
\centering
\caption{Block timing by transaction load}
\label{tab:timing-util}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Network State} & \textbf{Tx Rate} & \textbf{Block Time} \\
\midrule
Idle & $<$0.2 tx/s & 40s \\
Low usage & $\geq$0.2 tx/s & 20s \\
Medium usage & $\geq$1 tx/s & 10s \\
High usage & $\geq$5 tx/s & 5s \\
Very high usage & $\geq$20 tx/s & 3s \\
\bottomrule
\end{tabular}
\end{table}

The 5-second level is the \emph{monetary reference}: all emission and
inflation figures in this section are computed at 5\,s blocks
($N_{\text{year}} = 6{,}307{,}200$), the pace the network holds under
sustained high load. Slower levels produce proportionally fewer blocks per
year and therefore lower effective emission (Section~\ref{sec:dynamic-timing},
Emission Bounds); the 3-second floor is reached only under transient
maximum demand.

\input{figures/dynamic-timing}

\subsubsection{Economic Implications}

Dynamic timing creates feedback loops:

\textbf{Low utilization}:
\begin{itemize}
  \item Longer block times $\rightarrow$ fewer blocks per hour
  \item Fewer blocks $\rightarrow$ less emission
  \item Less emission $\rightarrow$ reduced inflation when currency is
    unused
\end{itemize}

\textbf{High utilization}:
\begin{itemize}
  \item Shorter block times $\rightarrow$ more blocks per hour
  \item More blocks $\rightarrow$ higher throughput
  \item Higher throughput $\rightarrow$ more rewards when currency is
    actively used
\end{itemize}

This aligns miner incentives with network utility.

\subsubsection{Emission Bounds}

Because the tail reward is fixed per block (at $\approx$2.42 BTH/block gross
for the $\approx$611M tail-onset supply), slower blocks produce fewer blocks
per year and therefore proportionally lower annual emission. The effective
annual tail emission thus varies with utilization:

\begin{table}[h]
\centering
\caption{Tail emission by load level (at $\approx$611M tail-onset supply)}
\label{tab:emission-util}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Load Level} & \textbf{Blocks/Year} & \textbf{Gross Emission/Year} \\
\midrule
Idle (40s)   & 788,400   & 1.91M BTH \\
Medium (10s) & 3,153,600 & 7.63M BTH \\
High, reference (5s) & 6,307,200 & 15.28M BTH \\
\bottomrule
\end{tabular}
\end{table}

At the 5\,s monetary reference this is the 2.5\% gross (2\% net) tail target;
sustained slower operation acts as a natural inflation dampener, holding
emission below the 2\% net rate. The 3-second floor (transient maximum demand)
would raise the block count to 10.5M/year, briefly exceeding the reference
target until difficulty and load re-equilibrate.

\subsection{Fee Economics}

\subsubsection{Base Fee}

The minimum fee per byte is:

\begin{equation}
f_{\text{base}} = 1 \text{ pico-BTH per byte}
\end{equation}

(\Botho's smallest unit is the picocredit, $1\text{ BTH} = 10^{12}$ pico-BTH.)
For a typical 4 KB transaction:
\begin{equation}
f_{\text{min}} = 4{,}000 \times 1 = 4{,}000 \text{ pico-BTH} = 4 \times 10^{-9} \text{ BTH}
\end{equation}

The 1\,pico-BTH/byte figure is the congestion \emph{floor}. Under load the
dynamic base rises multiplicatively (a congestion response bounded by a
$100\times$ safety clamp, i.e.\ $100$\,pico-BTH/byte), so the enforced
minimum is
\begin{equation}
f_{\min} = \underbrace{b_{\text{dyn}}(\text{congestion})}_{1\ldots100\text{ pico/byte}}
           \times \text{size} \times \text{cluster\_factor}
           \;+\; d(\vec{t}, v),
\end{equation}
where $d(\vec{t}, v)$ is the \emph{cluster demurrage} stock-level term
(Section~\ref{sec:cluster-tags}): a per-transfer charge proportional to the
spent value, the annual demurrage rate, and the coins' cluster factor, so
idle high-factor wealth carries a small holding cost that circulating
low-factor coins avoid. Demurrage is exempt for factor-1 coins.

\subsubsection{Progressive Fee Multiplier}

Transactions pay additional fees based on cluster wealth (see
Section~\ref{sec:cluster-tags}):

\begin{equation}
f_{\text{total}} = f_{\text{base}} \times \text{size} \times \text{cluster\_factor}
\end{equation}

The cluster factor ranges from 1.0$\times$ (bottom 50\% wealth) to
6.0$\times$ (top 1\% wealth).

\subsubsection{Demurrage Class-Transition Charge}
\label{sec:demurrage-class-transition}

Ordinary demurrage accrues \emph{over holding time} and is paid on each spend.
On its own this leaves a leak: a wealthy coin can be spent while young ---
before demurrage has accrued (\texttt{elapsed}~$\approx 0$) --- into a
background (factor-1) output, permanently escaping its class's future holding
cost at near-zero cost. \Botho\ closes this by charging, on every spend,
\begin{equation}
d_{\text{spend}} = \max\!\big(\underbrace{d_{\text{accrued}}}_{\text{time-based}},\;
\underbrace{d_{\text{cap}}}_{\text{capitalized reset}}\big),
\end{equation}
where $d_{\text{accrued}}$ is the ordinary accrued demurrage and the
\emph{capitalized reset charge} $d_{\text{cap}}$ prices a permanent
\emph{class downgrade} --- the case where a spend declares an output factor
\emph{below} the input's ring-implied class floor (the factor the spender
cannot rewrite). It equals the net future demurrage escaped by the downgrade,
capitalized over a settlement horizon $H = 5$ years
($\texttt{SETTLEMENT\_HORIZON\_BLOCKS} = 31{,}536{,}000$ blocks at the 5\,s
reference):
\begin{equation}
d_{\text{cap}} = d(\text{value}, \phi_{\text{floor}}, H) - d(\text{value}, \phi_{\text{out}}, H)
\quad\text{if } \phi_{\text{out}} < \phi_{\text{floor}},\; \text{else } 0.
\end{equation}
A genuine background$\to$background spend floors at 1$\times$ on both sides, so
$d_{\text{cap}} = 0$ and honest commerce is never touched; only an actual
downgrade relative to the ring floor is priced. The charge folds into the
consensus fee floor and thus flows, via the 80/20 split, into the same
lottery redistribution pool (ADR~0009; issue~\#925).

Because the floor is an order statistic over the ring's implied factors
weighted by input value, a single high-value decoy in an honest background spender's
ring could spuriously lift the floor above 1$\times$ and trip the charge. Decoy
selection therefore draws only ring members whose cluster factor lies within a
\textbf{$\pm 10\%$ factor-similarity band} of the real input's factor
(\texttt{FACTOR\_SIMILARITY\_SPREAD}), so a background spender draws only
near-1$\times$ decoys and the ring floor stays exact.

A holder may also \emph{opt in} to a \textbf{demurrage-settlement operation}
(issue~\#831): a transaction declaring all-background (factor-1) outputs pays
the same capitalized reset charge --- evaluated from the ring-floored input
class down to 1$\times$ over the same horizon $H$ --- to permanently reclassify
a wealthy coin to factor-1 and make it wrap-eligible for the bridge on-ramp
(Section~\ref{sec:bridge}). Both doors (the implicit \#925 downgrade charge and
the explicit \#831 settlement charge) read the \emph{one} shared horizon $H$
and reuse the same charge formula, so neither is a cheaper route out of the
demurrage regime than the other.

\subsubsection{Lottery-Based Fee Redistribution}

Transaction fees are split between a redistribution pool and burning:

\begin{itemize}
  \item \textbf{80\% to the lottery pool}: Added to a persistent,
    cumulative pool and distributed to 4 UTXOs per block via a verifiable,
    value-free uniform lottery over a circulation window (\emph{Path C};
    ADR~0009)
  \item \textbf{20\% burned}: Permanently removed from supply
\end{itemize}

The lottery pool is not fee-funded alone. Each block also routes a
height-scheduled \emph{emission share} of the block reward into the same
pool: the share ramps from 0 at genesis by 10 percentage points per halving
epoch to a 50\% cap (the miner keeps the remainder). Emission-routing is what
lets redistribution reach idle wealth that never transacts --- transaction
fees are a consumption tax and cannot, by themselves, touch coins that sit
still (Section~\ref{sec:economic-modeling}). The pool balance is cumulative
(carried over across blocks) and each block pays out
\begin{equation}
\text{payout} = \min\!\big(\underbrace{\text{carryover} + \text{emission\_share} + 0.8\,f_{\text{block}}}_{\text{available}},\; R\big),
\qquad
R = \min\!\big(\rho \cdot f_{\text{lott}},\; R(h)\big),
\end{equation}
where the \emph{reward cap} $R$ is the smaller of the Path~C Sybil cap
$\rho \cdot f_{\text{lott}}$ ($\rho$ = the count of in-window eligible
outputs, $f_{\text{lott}} = 0.25$ BTH the lottery base fee) and one block
reward $R(h)$. Any surplus \emph{carries forward} to later blocks rather than
being burned, so $\approx$100\% of collected fees are eventually redistributed.
The block-reward term $R(h)$ is the anti-grinding bound
(Section~\ref{sec:security}); the $\rho \cdot f_{\text{lott}}$ term is the
Sybil-neutrality bound (below).

This creates:

\begin{itemize}
  \item \textbf{Value-free selection}: Winners are drawn \emph{uniformly}
    over the circulation window --- no UTXO value and no cluster factor enters
    the draw, so the lottery reads only public chain state and remains
    compatible with confidential amounts (ADR~0009)
  \item \textbf{Reward-cap net-zero (Sybil-neutrality)}: A holder that splits
    a position into many outputs to buy more tickets also raises the reward
    cap $R = \rho \cdot f_{\text{lott}}$ by exactly one base fee per extra
    eligible output. It therefore wins back precisely its base-fee outlay ---
    realized net gain is \emph{zero} at every split factor, and strictly
    negative whenever the fee pool is thin. Sybil resistance comes from the
    cap, not from a value-based weight formula
  \item \textbf{Progressive intake}: Redistribution progressivity now lives
    on the \emph{source} side --- the demurrage and progressive-fee terms
    that fill the pool are value- and factor-dependent, while the payout draw
    is Sybil-neutral
  \item \textbf{Deflationary pressure}: 20\% burn creates supply reduction
  \item \textbf{Anti-MEV}: No miner incentive to reorder transactions
  \item \textbf{Alignment}: Miner income comes from block rewards only
\end{itemize}

Earlier designs drew winners with weight
$w_i = v_i \times (\phi_{\max} - \phi_i + 1)$ --- proportional to UTXO value
and tilted by the inverse cluster factor. That weight is \emph{split-invariant}
(splitting conserves value hence weight), but it reads the UTXO value directly,
which is incompatible with the confidential amounts of ADR~0006/0009, and
adversarial simulation (the \texttt{cluster-tax} lottery-selection sweep;
Section~\ref{sec:economic-modeling}) showed the value term dominates the
factor tilt in realistic populations, routing much of the pool
\emph{back} to whales. Path~C (ADR~0009,
\texttt{docs/research/ct-compatible-lottery-selection.md}) retires the value
tilt entirely: the payout is a uniform draw whose split-resistance is the
endogenous reward cap, not the weight. Because consensus must count a
splitter's \emph{own} outputs in $\rho$, the cap rises in lockstep with the
tickets bought, pinning the attacker's realized gain at zero --- a
regime-independent Sybil bound obtained from public signals alone, with no
zero-knowledge value proof. In honest populations this is strictly more
equalizing than the retired tilt.

\input{figures/fee-flow}

\subsubsection{Lottery Mechanism}

The draw runs once \emph{per block}, over the block's accumulated pool:

\begin{enumerate}
  \item Compute the fee split: burn $0.2\,f_{\text{block}}$; add
    $0.8\,f_{\text{block}}$ plus the emission share to the cumulative pool
  \item Filter eligible candidates: UTXOs at least 720 blocks old, worth at
    least $10^{6}$ pico-BTH (1 micro-BTH), and \emph{spent-and-recreated
    within the last $N = 10{,}000$-block circulation window} --- idle wealth
    self-excludes and the base fee is the only value gate. Let $\rho$ be the
    resulting count of eligible outputs
  \item Set the reward cap $R = \min(\rho \cdot f_{\text{lott}},\, R(h))$,
    with $f_{\text{lott}} = 0.25$ BTH the lottery base fee and $R(h)$ one
    block reward; pay out $\min(\text{available},\, R)$ this block and carry
    any surplus forward
  \item Derive the seed from the \emph{committed previous block hash} and
    height,
    \begin{equation}
    s = \Hash(\text{``LOTTERY\_SEED\_V1''} \,\|\, \text{prev\_block\_hash} \,\|\, h),
    \end{equation}
    and select 4 winners \emph{uniformly} without replacement --- every
    eligible UTXO has equal probability, with no value or cluster-factor
    weight
  \item Split the capped payout equally among the 4 winners
\end{enumerate}

Because the seed depends only on the already-externalized previous block, a
block proposer cannot bias the draw through transaction selection or ordering
of the \emph{current} block. Two bounds complete the security argument: the
per-block reward cap makes seed-grinding unprofitable
(Section~\ref{sec:security}, Lottery Grinding), and the Sybil cap
$\rho \cdot f_{\text{lott}}$ makes ticket-splitting net-zero (the reward-cap
property above).

\subsubsection{Effective Inflation}

The effective inflation rate becomes:

\begin{equation}
\text{inflation}_{\text{effective}} = \frac{\text{emission} - \text{fees\_burned}}{\text{supply}}
\end{equation}

where $\text{fees\_burned} = 0.2 \times \text{total\_fees}$. Under high
utilization with progressive fees, burning may exceed tail emission,
creating net deflation.

\subsubsection{Fee Flow Analysis}

\begin{table}[h]
\centering
\caption{Fee flow scenarios (1M daily transactions)}
\label{tab:fee-flow}
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Wealth Distribution} & \textbf{Daily Fees} & \textbf{Redistributed} & \textbf{Burned} \\
\midrule
Equal (factor 1.0) & 24.7 mBTH & 19.8 mBTH & 4.9 mBTH \\
Current inequality & 74.1 mBTH & 59.3 mBTH & 14.8 mBTH \\
High inequality & 148.2 mBTH & 118.6 mBTH & 29.6 mBTH \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Assumptions and derivation}: the per-transaction figure is the
congestion-floor consensus minimum for a standard transfer, computed directly
from the fee formula (\texttt{minimum\_fee\_dynamic\_with\_outputs},
\texttt{botho/src/ledger/store.rs}). Take a standard 2-output transfer
(payment + change) from a factor-1 cluster. Under the universal ML-KEM-768
stealth format (Section~\ref{sec:pq-stealth}) every output carries a mandatory
1{,}088-byte encapsulation ciphertext, so a $\approx$4{,}000-byte transaction
body grows by $2\times1{,}088 = 2{,}176$ bytes to
$\text{size} \approx 6{,}176$ bytes. At the congestion floor
$b = 1$\,pico-BTH/byte:
\begin{equation}
f_{\text{tx}} = \underbrace{b}_{1} \times \underbrace{\text{size}}_{6{,}176}
  \times \underbrace{\text{cluster\_factor}}_{1.0}
  \times \underbrace{\text{output\_penalty}}_{2^{2} = 4}
  = 24{,}704 \text{ pico-BTH} = 2.47\times10^{-8}\,\text{BTH},
\end{equation}
i.e.\ $\approx$24.7\,nano-BTH per transaction (the 2-output superlinear penalty
$\min(2,10)^2 = 4$ is part of the enforced minimum). Over 1M
transactions/day the factor-1 aggregate is
$10^{6}\times24{,}704\text{ pico-BTH} = 2.47\times10^{10}\text{ pico-BTH}
 = 2.47\times10^{-2}\,\text{BTH} \approx 24.7\,\text{mBTH}$. The remaining rows scale
the aggregate by each distribution's mean cluster factor ($\approx3\times$ and
$\approx6\times$, within the 1--6 range). These are floor-level flow
magnitudes; a rising dynamic base (up to the $100\times$ clamp) or larger,
multi-output transactions raise them proportionally.

\subsection{Minter Incentives}

\subsubsection{Block Reward Composition}

Minters receive only the block reward:

\begin{equation}
\text{minter\_income} = R(h) \text{ (no fees)}
\end{equation}

This simplifies minting economics and eliminates fee-based attacks.

\subsubsection{Profitability Analysis}

Minting profitability depends on:
\begin{itemize}
  \item Hardware cost (one-time)
  \item Electricity cost (ongoing)
  \item Block reward value
  \item Network hashrate (competition)
\end{itemize}

The tail emission ensures perpetual profitability for efficient miners,
unlike Bitcoin where fee-only income creates uncertainty.

\subsubsection{Decentralization Incentives}

Several mechanisms encourage mining decentralization:

\begin{itemize}
  \item \textbf{CPU-friendly PoW}: Algorithm resists ASIC development
    (RandomX-based~\cite{randomx})
  \item \textbf{Solo mining viable}: Tail emission ensures consistent
    income even at low hashrate
  \item \textbf{No economy of scale}: Linear scaling of rewards with
    hashpower
\end{itemize}

\subsection{Long-Term Sustainability}

\subsubsection{Security Budget}

The perpetual security budget (in BTH terms) is:

\begin{equation}
\text{security\_budget} = R_{\text{tail}} \times \text{blocks\_per\_year}
\end{equation}

At the $\approx$611M tail-onset supply this ranges from $\approx$1.9M BTH (idle)
to $\approx$15.3M BTH (maximum utilization) annually, and grows with supply
thereafter.

\subsubsection{Comparison with Fixed Supply}

\begin{table}[h]
\centering
\caption{Security model comparison: fixed supply vs. tail emission}
\label{tab:supply-security-comparison}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Aspect} & \textbf{Fixed Supply} & \textbf{Tail Emission} \\
\midrule
Long-term miner income & Fees only & Rewards + fees \\
Security if fees low & Degraded & Maintained \\
Inflation & 0\% & $\sim$2--3\% \\
Predictability & High & High \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Wealth Tax Equivalence}

Tail emission functions as a mild wealth tax:

\begin{equation}
\text{effective\_tax} = \frac{\text{tail\_emission}}{\text{total\_supply}} \approx 2\%
\end{equation}

This encourages circulation: holding idle currency means gradual dilution,
while active participation maintains relative wealth.

\subsection{Monetary Constants}

\begin{table}[h]
\centering
\caption{Monetary policy constants}
\label{tab:constants}
\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Description} \\
\midrule
Initial reward & 50 BTH & Block reward at genesis \\
Halving period & 6,307,200 blocks & $\approx$ 1 year (at 5s blocks) \\
Halving count & 5 & Time-to-tail $\approx$ 5 years \\
Phase-1 supply & $\approx$611M BTH & Geometric sum of 5 halvings \\
Tail inflation & 2\% net/yr & Supply-dependent, perpetual \\
Tail reward & $\approx$1.94 BTH/block & Net, at 611M supply (grows with supply) \\
Min block time & 3 seconds & Consensus floor (very high load) \\
Reference block time & 5 seconds & Monetary-calculation reference \\
Max block time & 40 seconds & Idle ceiling \\
Base fee rate & 1 pico-BTH/byte & Congestion floor (dynamic, $\leq$100$\times$) \\
Lottery winners/block & 4 & Value-free uniform draw (Path C) \\
Lottery circulation window & 10,000 blocks & Eligibility horizon ($\approx$14\,h) \\
Lottery base fee & 0.25 BTH & Reward cap $R = \min(\rho f_{\text{lott}}, R(h))$ \\
Emission-to-lottery cap & 50\% of reward & Ramps 10\%/halving epoch \\
Settlement horizon & 5 years & Capitalized reset / class-downgrade charge \\
Decoy factor band & $\pm 10\%$ & Keeps ring floor exact on 1$\times$ spends \\
Max cluster factor & 6.0$\times$ & Top 1\% wealth multiplier \\
Decimals & 12 & Smallest unit: 1 picocredit \\
\bottomrule
\end{tabular}
\end{table}

